Distinguishing multiple-merger from Kingman coalescence using two-site frequency spectra

### 2018-10-07 manuscript/manuscript.tex
Maryn suggestions (part 2).

This transformation of the 2-SFS has several useful properties.
First, because it is based on the minor allele frequencies, we may compute $\fpmi$ from data without knowing the ancestral allele.
Moreover, the results of \cite{Fu1995} imply that $\fpmi(i,j) < 0$ for $i\neq j$ for non-recombining sites under the time-homogeneous Kingman coalescent.

The primary benefit of computing $\fpmi$ from the 2-SFS is that it normalizes for distortions in the coalescent time distribution.
Population-size variation distorts the SFS by changing the distribution of waiting times.
However, the time-inhomogeneous Kingman coalescent generates the same distribution of tree topologies as the time-homogeneous coalescent.
In contrast, multiple mergers alter both the coalescent time distribution and the distribution of topologies.
These two effects are convolved in the SFS, making it difficult to distinguish between population growth and multiple mergers.
Our hope is that the $\fpmi$ reflects additional information about the distribution of genealogies, once distortions in coalescent times are accounted for.
That is, we expect $\fpmi$ to be sensitive to multiple mergers, but relatively unaffected by changes in the coalescent rate.

Furthermore, the same normalization renders $\fpmi$ insensitive to ascertainment bias.
Variant detection methods typically compare sampled sequences to a reference sequence and ``call'' a variant site when there is sufficient evidence that at least one sample differs from the reference at that site.
As a result, the ascertainment probability, $p_i$, of a mutation is an increasing function of its true allele count $i$.
This effect distorts the expected ascertained SFS ($\E{\eta_i}_{\text{asc}} = \mu \E{\tau_i} p_i$) toward high-frequency mutations.
However, provided that variant detection at each site is approximately independent, the primary effect of ascertainment bias in \eq{eq:fpmi} is to multiply the numerator and denominator by a common factor of $p_i p_j$.
Because these factors cancel, $\fpmi$ is less sensitive than the raw SFS and 2-SFS to ascertainment effects.
% A general measure of the dependence between a pair of random variables is the \textit{mutual information} \autocite{CoverThomas1991}.
% The mutual information between random variables $X$ and $Y$ is a functional of their joint probability mass function, $p(x,y)$, defined as
% \begin{align}
%     I[p(x,y)] = \sum_{x,y} p(x,y) \log \frac{p(x,y)}{p(x)p(y)},
%     \label{eq:mutual_information}
% \end{align}
% where $p(x) = \sum_y p(x,y)$ and vice versa.
% Equivalently, the mutual information is the Kullback-Leibler divergence between the true joint distribution of $X$ and $Y$ and the joint distribution of two independent random variables with the same marginal distributions as $X$ and $Y$ \autocite{CoverThomas1991}.
% Thus, mutual information captures generic, possibly nonlinear, dependence between $X$ and $Y$.

@@

With finite data, estimates of the 2-SFS will be noisy.
This is particularly true for $i,j\gg1$ because $\E{\eta_{ij}}$ decays like $(ij)^{-1}$ for the standard Kingman \autocite{Fu1995} and faster with growth or multiple mergers.
We show in Results that a positive association between mutations with high minor allele counts and mutations with low minor allele counts is a signature of multiple mergers.
This suggests using a binned form of the SFS and 2-SFS:
\begin{align}
    \eta_{\text{lo}}(i_c) &= \sum_{i=1}^{i_c-1} \eta_i \\
    \eta_{\text{hi}}(i_c) &= \sum_{i=i_c}^{\floor{n/2}} \eta_i \\
    \eta_{\text{hi,lo}}(d; i_c) &= \sum_{i=1}^{i_c-1} \sum_{j=i_c}^{\floor{n/2}} \eta_{ij}(d),
\end{align}
where $i_c$ is an arbitrary cutoff between high and low minor allele frequency.
Binning allows us to estimate the 2-SFS stably for large sample sizes, because we may adjust $i_c$ to ensure a large number of sites in both the high--minor allele count and low--minor allele count bins.
It is possible to define less drastic coarse-graining schemes that strike a balance between sampling noise and preserving more detailed information about allele frequencies.

We can also compute the pointwise mutual information in this binned distribution as
\begin{align}
    \hilopmi(d; i_c) = \log \frac
                                {\E{\eta_{\text{hi,lo}}(d; i_c)}}
                                {\E{\eta_{\text{lo}}(i_c)}  \E{\eta_{\text{hi}}(i_c)}}.
\end{align}
One could similarly calculate five other pointwise mutual information statistics from the binning 2-SFS (e.g., the PMI between monomorphic sites and sites with high minor allele counts), but we will not use these statistics here.

\subsection*{Weighted $\fpmi$}
For plotting purposes, we will also use a weighted version of the $\fpmi$:
\begin{align}
    \wfpmi(i,j:d) = \frac{\E{\eta_{ij}}}{\mu^2 \E{T_2}^2} \fpmi(i,j;d),
    \label{eq:wfpmi}
\end{align}
where $T_2$ is the coalescence time for a sample of size two.
The numerator of the weighting factor in \eq{eq:wfpmi} serves to emphasize the most common pairs of minor allele counts.
The denominator, which is proportional to the square of the expected pairwise diversity, $\Pi$, ensures that $\wfpmi$ is invariant to changes in the mutation rate and average pairwise coalescence time.

@@

Functions for computing the branch-length moments are implimented in \texttt{python} using the numerical package \texttt{numpy} and available [GITHUB].
The Kingman coalescent code currently can compute moments for exponentially growing and two-epoch piecewise constant models, but could be extended to allow for other models.
The formulas of \cite{ZivkovicWiehe2008} exhibit numerical instability related to the instability of \cite{GriffithsTavare1994}.
They are therefore only practical for sample sizes up to $n \approx 40$.
The recursion of \cite{BirknerEtAl2013} is $\mathcal{O}(n^4)$ and is also only practical for samples up to $n \approx 50$.

@@

For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with length $d$ and per basepair recombination rate $r$ resulting in a per locus total recombination rate $d\,r$.
(See SI for parameter combinations.)
We chose values of $d\,r$ so that the realized values of $d\,r \E{T_2}$ varied over several orders of magnitude.
For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at each end of the loci.
We then calculated the expectations $\{\E{\tau_i}\}$ and $\{\E{\tau_{ij}}\}$ by averaging over independent loci.
These expectations allow us to calculate all of the 2-SFS statistics defined above.

@@

We analyzed sequence data from the DPGP3 data set, which consists of haploid consensus sequences from $\sim 200$ flies, obtained via the haploid embryo method of \cite{LangleyEtAl2011}.
The SNP calls underlying these sequences have been subjected to a variety of quality filters described in \cite{LackEtAl2015}.
We obtained the DPGP3 consensus sequence files version 1.1 from \url{www.johnpool.net/genomes.html}.
These files contain the sequence alignments of all flies in the sample on all chromosome arms.
We also downloaded the Nov. 3, 2016 spreadsheet of inversions available at the link above.
For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by taking the first 100 remaining samples in alphanumeric order by sample name.
(As a result, the data for each chromosome arm is from a slightly different subset of the individuals.)

We calculated the average pairwise diversity, $\Pi$, as a function of position for each autosomal chromosome arm (\fig{fig:dpgp_pi}).
Pairwise diversity is high in the middle of each chromosome arm and lower near the centromeres and telomeres.
Our modeling---and coalescent-based demographic inference methods in general---assumes that the distribution of gene genealogies is homogeneous along the chromosome.
Therefore, we selected a 13-16~MB ``central'' region of each arm with relatively homogeneous values of $\Pi$ for further analysis.
The boundaries of these central regions are given in Table~\ref{tab:central_regions}.

In order to ensure that the segregating mutations reflect true genetic diversity and not variation in calling errors, we excluded sites where fewer than 90 of the 100 samples have genotype calls.
This leaves over 90\% of all sites and does not substantially alter the fraction of sites that are polymorphic in the sample (Table~\ref{tab:called_sites})

We fit a demographic model to the folded SFS of fourfold degenerate sites for each chromosome arm separately using \texttt{fastNeutrino} \autocite{BhaskarEtAl2015}.
Following \cite{RagsdaleGutenkunst2017} we fit a three-epoch piecewise constant--$N$ model:
\begin{align}
    N(t) = \begin{cases}
                N_1 & 0 \leq t \leq t_1 \\
                N_2 & t_1 <  t \leq t_2 \\
                N_{\text{anc}} & t > t_2,
            \end{cases}
    \label{eq:piecewise}
\end{align}
estimating both change-points ($t_1$ < $t_2$) and both population sizes ($N_1$, $N_2$).
We specified the ancient population size $N_{\text{anc}} = 3\times 10^5$, as in \cite{RagsdaleGutenkunst2017}.
For all four chromosomes, \texttt{fastNeutrino} inferred similar population growth.
Fitted parameters are presented in Table~\ref{tab:DPGP_params}.
We simulated the SFS and 2-SFS under the fitted parameters using \texttt{msprime}.

In addition to the average SFS used to fit the model, we computed the average 2-SFS for pairs of sites at distances distances between 3~bp and 5~Kb.
Because we are using four-fold degenerate sites, we only computed the 2-SFS for distances that are multiples of 3.
For comparison between data and simulations, we scale the distances in basepairs by a critical distance $d_c = (r\E{T_2})^{-1}$.
We used a genome-wide recombination rate of $r = 2 \times 10^{-8}$ per basepair per generation \autocite{ComeronEtAl2012}.
We estimated $\E{T_2}$ by $\Pi / 2 \mu$.
For $\Pi$, we used the average pairwise diversity at fourfold degenerate sites in the central region of each chromosome arm.
For $\mu$, we used a genome-wide mutation rate of $3 \times 10^{-9}$ per basepair per generation \autocite{KeightleyEtAl2014}.
These estimates are not precise, but only serve to scale genetic distances to the correct order of magnitude.

@@

\begin{figure}
\centering
\includegraphics[scale=1]{figures/wfpmi_no_recombination.pdf}
\caption{Weighted frequency pointwise mutual information for three coalescent models: the time-homogeneous Kingman coalescent, the Kingman coalescent with exponential growth ($g=4$), and the beta coalescent ($\alpha=1.45$). For all models $n=39$. Growth and $\alpha$ parameters were chosen to generate average SFS with similar Tajima's D. The diagonal $i=j$ is masked. \label{fig:nonrecombining_pmi}}
\end{figure}

As a first example, \fig{fig:nonrecombining_pmi} shows the $\wfpmi$ for constant-$N$ Kingman coalescent; the Kingman coalescent with exponential growth, $N(t)=N_0 \exp(-g \frac{t}{N_0})$; and a beta coalescent intermediate between the Kingman and Bolthausen-Sznitman coalescents.
Exponential growth with $g=4$ and the beta coalescent with $\alpha=1.45$ both generate similar substantial distortions in the SFS $D = -0.4$.
However, exponential growth does not qualitatively change the $\fpmi$.
In particular, $\fpmi_{i,j} < 0$ for all $i \neq j$, just as in the constant-$N$ Kingman coalescent.
On the other hand, the beta coalescent generates positive $\fpmi$.
The effect of multiple mergers is strongest on $\fpmi$ between high and low minor allele counts.
This justifies binning the SFS and 2-SFS into high-- and low--minor allele count bins as defined above.

To generalize this finding, we computed the binned $\hilopmi$ between singletons and non-singletons ($i_c = 1$) for a range of $g$ and $\alpha$.
\Fig{fig:fpmi_vs_tajimasd} shows that as the population growth rate increases, distorting the SFS, $\hilopmi$ increases relative to the constant-$N$ Kingman, but plateaus at a negative value.
On the other hand, beta coalescents that generate similar distortions in the SFS, generate larger changes in $\hilopmi$, including positive values.
Thus, the 2-SFS in general, and $\hilopmi$ in particular, are capable of capturing the effects of multiple mergers beyond the distortions in branch lengths.

@@

We have shown in the previous sections that $\fpmi$ between pairs of non-recombining sites can discriminate between the Kingman coalescent with population growth and the beta coalescent.
However, most demographic inference is performed on regions of the genome with non-zero recombination rates.
The advantage of the $\fpmi$ approach is that it can be computed on a genomic scale, as with the SFS.
Therefore, it is important to assess the robustness of our approach to recombination between sites.

To measure the relationship between $\fpmi$ and recombination, we ran coalescent simulations using a version of the program \texttt{msprime} \cite{KelleherEtAl2016} modified to allow for multiple mergers (See Methods).
In particular, we simulated a model where the sizes of coalescent events are drawn from the distribution specified by the beta coalescent, as in our numerical calculations above.
In this model, marginal genealogies will follow the beta coalescent distribution, and the average SFS will be given by the formula in \cite{BirknerEtAl2013}.
We also simulated data from two models of population growth: exponential growth and a two-epoch piecewise-constant model.

@@

\begin{figure}
\centering
\includegraphics[scale=1]{figures/hilopmi_vs_distance.pdf}
\caption{Pointwise mutual information between high-- and low--minor allele count mutations ($\hilopmi$) versus genetic distance. Each curve represents a different combination of coalescent parameters. The first three rows show coalescent simulations in \texttt{msprime}. Exponential growth rate $g$ varies logarithmically from 0.01 to 8.0 in coalescent time units. Piecewise constant change-points, range from 0.01 to 1 in coalescent time units, and the ratio of ancestral to modern population size ranges from 0.01 to 0.2. The fourth row shows forward-time \texttt{SLiM} simulations with selective sweeps at linked sites (beneficial mutation fraction $f_{\text{sel}} \in \{10^{-6}, 10^{-5}, 10^{-4}, 10^{-3}\}$, see Methods). Lines are colored by the distortion in the SFS, as measured by Tajima's D. \label{fig:hilo_vs_d}}
\end{figure}

\Fig{fig:hilo_vs_d} shows $\hilopmi(d,i_c)$ in both models of growth and the beta coalescent, for a range of $d\,r \E{T_2}$ and three different choices of $i_c$.
Each curve represents a particular parameter combination and coalescent model and is colored by the distortion in the average SFS, which is independent of the recombination rate.
At low recombination rates, $\hilopmi$ may be greater in growing populations than in constant-$N$ populations (dotted lines), but is always less than zero, consistent with the results for non-recombining sites.
When $d\,r \E{T_2}\geq 1$, $\hilopmi$ may be slightly positive for large $i_c$, but is smaller in growing populations than in the constant-$N$ Kingman model.
In all Kingman models, $\hilopmi$ decays to zero for $d\,r \E{T_2}\gg 1$.

@@

There is one other salient feature of \fig{fig:hilo_vs_d}: $\hilopmi$ does not decay to zero at long distances in the beta coalescent.
This is related to the results of \cite{EldonWakeley2006}, who showed that a model with ``jackpot'' reproductive events can generate infinite-range linkage disequilibrium.
They showed that different scalings of rates of mutation, pairwise coalescence, multiple merger coalescence, and recombination lead to different behaviors of diversity and linkage disequilibrium.
Our implementation of the beta coalescent model with recombination corresponds to a particular scaling limit where recombination does not have time to decorrelate trees during multiple mergers events, even at infinite genetic distances.
We do not expect this behavior to be universal in multiple mergers coalescents with recombination.

@@

To test this proposition, we performed forward-time simulations of recurring selective sweeps using the software \texttt{SLiM} \autocite{Messer2013}.
In these simulations, we simulated individual chromosomes with homogenous recombination and two types of mutations: neutral and beneficial with fixed selection coefficient $s$.
Both types of mutations occurred at random, uniformly distributed across the length of the chromosome.
By varying the recombination rate, mutation rates, population size, and selection coefficient, we were able to vary the rate of selective sweeps linked to neutral sites.

The bottom row of \fig{fig:hilo_vs_d} shows the results of these simulations.
For intermediate genetic distances, $d\,r \E{T_2} \sim 1$, the effects of linked sweeps are qualitatively similar to the effects of the beta coalescent.
That is, when sweeps are sufficiently frequent to distort the SFS, as measured by Tajima's D, they also increase $\hilopmi$.
The primary difference between the forward-time sweeps and beta coalescent simulations, is that the distortions caused by sweeps decay to zero for $d\,r \E{T_2} \gg 1$.
This decay is reasonable because the effect of a single selective sweep on the genealogy is localized around the position of beneficial mutation.

@@

Our results above show that $\fpmi$ and its binned analog, $\hilopmi$ are useful for distinguishing population growth from the effects of multiple mergers.
This is true even when the population growth generates similar distortions in the SFS.
We therefore propose the following model-checking procedure for demographic inference methods
\begin{enumerate}[label=(\roman*), before=\unskip{: }]
    \item Fit a demographic model to data.
    \item Simulate genealogies under the fitted model (using \texttt{msprime} or other coalescent simulator). Calculate the $\fpmi(d)$ and $\hilopmi(d;i_c)$ predicted by the model.
    \item Calculate $\fpmi(d)$ and $\hilopmi(d;i_c)$ from the data.
    \item Compare true to predicted statistics. Evaluate model fit.
\end{enumerate}
This procedure checks whether the model is consistent with a dimension the data that was not used in fitting the model.
Inconsistency suggests that the inferred $N(t)$ may be an artifact of natural selection, skewed offspring distributions, etc., rather than reflecting the true historical population size.

In this section, we illustrate the procedure just outlined by using genomic diversity data from the \textit{Drosophila melanogaster} DGPG3 panel \cite{LackEtAl2015}.
The DPGP3 data consists of haploid consensus sequences from $\sim$ 200 wild-caught flies from a Zambian population known to be relatively free of cosmopolitan admixture.
Recently, several groups have used the DPGP3 data to estimate the population-size history of \textit{D. melanogaster} \autocite{TerhorstEtAl2017,RagsdaleGutenkunst2017}.
On the other hand, it is widely believed that the genetic diversity of \textit{Drosophila} is strongly shaped by natural selection (e.g., \cite{ElyashivEtAl2016, GarudPetrov2016})
Thus, this data is a good candidate for demonstrating the utility of $\fpmi$ for assessing coalescent model fit.

>>>>

This transformation of the 2-SFS has several useful properties.
First, because $\fpmi$ is based on the minor allele frequencies, we may compute it without knowing the ancestral allele.
Second, the results of \cite{Fu1995} for the folded SFS imply that $\fpmi(i,j) < 0$ for $i\neq j$ for non-recombining sites under the time-homogeneous Kingman coalescent.
Finally, $\fpmi$ normalizes for distortions in the coalescence time distribution reflected in the single site SFS.
In the Kingman coalescent, population-size variation alters the coalescence time distribution but preserves the distribution of tree topologies.
We will show that, because $\fpmi$ normalizes the 2-SFS by the product of single site spectra, it is relatively insensitive to the distribution of coalescence times.
Thus, $\fpmi$ primarily reflects the distribution of topologies, which distinguishes multiple mergers from the Kingman coalescent.

Furthermore, the same normalization renders $\fpmi$ insensitive to ascertainment bias.
Variant detection methods typically compare sampled sequences to a reference sequence and ``call'' a variant site when there is sufficient evidence that at least one sample differs from the reference at that site.
As a result, the ascertainment probability, $p_i$, of a mutation is an increasing function of its true allele count $i$, skewing the expected ascertained SFS ($\E{\eta_i}_{\text{asc}} = \mu \E{\tau_i} p_i$) toward high-frequency mutations.
However, provided that variant detection at each site is approximately independent, the primary effect of ascertainment bias in \eq{eq:fpmi} is to multiply the numerator and denominator by a common factor of $p_i p_j$.
Because these factors cancel, $\fpmi$ is less sensitive than the raw SFS and 2-SFS to ascertainment effects.
% A general measure of the dependence between a pair of random variables is the \textit{mutual information} \autocite{CoverThomas1991}.
% The mutual information between random variables $X$ and $Y$ is a functional of their joint probability mass function, $p(x,y)$, defined as
% \begin{align}
%     I[p(x,y)] = \sum_{x,y} p(x,y) \log \frac{p(x,y)}{p(x)p(y)},
%     \label{eq:mutual_information}
% \end{align}
% where $p(x) = \sum_y p(x,y)$ and vice versa.
% Equivalently, the mutual information is the Kullback-Leibler divergence between the true joint distribution of $X$ and $Y$ and the joint distribution of two independent random variables with the same marginal distributions as $X$ and $Y$ \autocite{CoverThomas1991}.
% Thus, mutual information captures generic, possibly nonlinear, dependence between $X$ and $Y$.

@@

With finite data, estimates of the 2-SFS will be noisy.
This is particularly true for $i,j\gg1$ because $\E{\eta_{ij}}$ decays like $(ij)^{-1}$ for the standard Kingman \autocite{Fu1995} and faster with growth or multiple mergers.
We show in Results that a positive association between mutations with high minor allele counts and mutations with low minor allele counts is a signature of multiple mergers.
Thus, to detect multiple mergers, we use a binned form of the SFS and 2-SFS:
\begin{align}
    \eta_{\text{lo}}(i_c) &= \sum_{i=1}^{i_c-1} \eta_i \\
    \eta_{\text{hi}}(i_c) &= \sum_{i=i_c}^{\floor{n/2}} \eta_i \\
    \eta_{\text{hi,lo}}(d; i_c) &= \sum_{i=1}^{i_c-1} \sum_{j=i_c}^{\floor{n/2}} \eta_{ij}(d),
\end{align}
where $i_c$ is an arbitrary cutoff between high and low minor allele frequency.
Binning provides a stable estimate of the 2-SFS for large sample sizes, because we may adjust $i_c$ to ensure a large number of sites in both the high--minor allele count and low--minor allele count bins.
We note that a less coarse-grained binning scheme may strike a balance between sampling noise and preserving more detailed information about allele frequencies.

We compute the pointwise mutual information of the binned distribution as
\begin{align}
    \hilopmi(d; i_c) = \log \frac
                                {\E{\eta_{\text{hi,lo}}(d; i_c)}}
                                {\E{\eta_{\text{lo}}(i_c)}  \E{\eta_{\text{hi}}(i_c)}}.
\end{align}
While we will focus on $\hilopmi$, one could similarly calculate five other pointwise mutual information statistics from the binned 2-SFS (e.g., the PMI between monomorphic sites and sites with high minor allele counts).

\subsection*{Weighted $\fpmi$}
For plotting purposes, we use a weighted version of the $\fpmi$:
\begin{align}
    \wfpmi(i,j:d) = \frac{\E{\eta_{ij}}}{\mu^2 \E{T_2}^2} \fpmi(i,j;d),
    \label{eq:wfpmi}
\end{align}
where $T_2$ is the coalescence time for a sample of size two.
The numerator of the weighting factor in \eq{eq:wfpmi} emphasizes the most common pairs of minor allele counts.
The denominator, which is proportional to the square of the expected pairwise diversity, $\Pi$, ensures that $\wfpmi$ is invariant to changes in the mutation rate and average pairwise coalescence time.

@@

Functions for computing the branch-length moments are implimented in \texttt{python} using the numerical package \texttt{numpy} and available [GITHUB].
Our Kingman coalescent code can compute moments for exponentially growing and two-epoch piecewise constant models, but could be extended to allow for other models.
The formulas of \cite{ZivkovicWiehe2008} exhibit numerical instability related to the instability described in \cite{GriffithsTavare1994}.
Thus, these formulas are only practical for sample sizes up to $n \approx 40$.
The recursion of \cite{BirknerEtAl2013} is $\mathcal{O}(n^4)$ and is also only practical for samples up to $n \approx 50$.

@@

For each coalescent model, we simulated at least $10^4$ independent infinite-sites loci, each with length $d$ and per basepair recombination rate $r$ resulting in a per locus total recombination rate $d\,r$.
(See SI for parameter combinations.)
We chose values of $d\,r$ so that the realized values of $d\,r \E{T_2}$ varied over several orders of magnitude.
For each locus, we measured $\{\tau_i : i = 1,\ldots,n-1\}$ in the two genealogies at the ends of the locus.
We then calculated the expectations $\{\E{\tau_i}\}$ and $\{\E{\tau_{ij}}\}$ by averaging over independent loci.
These expectations allow us to calculate all of the 2-SFS statistics defined above.

@@

We analyzed sequence data from the DPGP3 data set, which consists of haploid consensus sequences from $\sim 200$ flies, obtained via the haploid embryo method of \cite{LangleyEtAl2011}.
The SNP calls that characterize these sequences were subjected to a variety of quality filters described in \cite{LackEtAl2015}.
We obtained the DPGP3 consensus sequence files version 1.1 from \url{www.johnpool.net/genomes.html}.
These files contain the sequence alignments of all flies in the sample on all chromosome arms.
We also downloaded the Nov. 3, 2016 spreadsheet of inversions available at the link above.
For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by selecting the first 100 remaining samples in alphanumeric order by sample name.
As a result, the data for each chromosome arm is from a slightly different subset of the individuals.

We calculated the average pairwise diversity, $\Pi$, as a function of position for each autosomal chromosome arm (\fig{fig:dpgp_pi}).
Pairwise diversity is high in the middle of each chromosome arm and lower near the centromeres and telomeres.
Our modeling---and coalescent-based demographic inference in general---assumes that the distribution of gene genealogies is homogeneous along the chromosome.
Therefore, we selected a 13-16~MB ``central'' region of each arm with relatively homogeneous values of $\Pi$ for further analysis.
The boundaries of these central regions are given in Table~\ref{tab:central_regions}.

In order to ensure that the segregating mutations reflect true genetic diversity and not variation in calling errors, we excluded sites with fewer than 90 of the 100 genotypes called.
This leaves over 90\% of all sites and without substantially altering the fraction of polymorphic sites (Table~\ref{tab:called_sites}).

We fit a demographic model to the folded SFS of fourfold degenerate sites for each chromosome arm separately using \texttt{fastNeutrino} \autocite{BhaskarEtAl2015}.
Following \cite{RagsdaleGutenkunst2017} we fit a three-epoch piecewise constant--$N$ model:
\begin{align}
    N(t) = \begin{cases}
                N_1 & 0 \leq t \leq t_1 \\
                N_2 & t_1 <  t \leq t_2 \\
                N_{\text{anc}} & t > t_2,
            \end{cases}
    \label{eq:piecewise}
\end{align}
estimating both change-points ($t_1 < t_2$) and population sizes ($N_1$, $N_2$).
We specified the ancient population size $N_{\text{anc}} = 3\times 10^5$, as in \cite{RagsdaleGutenkunst2017}.
For all four chromosomes, \texttt{fastNeutrino} inferred similar population growth.
Fitted parameters are presented in Table~\ref{tab:DPGP_params}.
We simulated the SFS and 2-SFS under the fitted parameters using \texttt{msprime}.

In addition to the average SFS used to fit the model, we computed the average 2-SFS for pairs of sites at distances distances between 3~bp and 5~Kb.
Because we are using fourfold degenerate sites, we only computed the 2-SFS for distances that are multiples of three.
For comparison between data and simulations, we scale the distances in basepairs by a critical distance $d_c = (r\E{T_2})^{-1}$.
We used a genome-wide recombination rate of $r = 2 \times 10^{-8}$ per basepair per generation \autocite{ComeronEtAl2012}.
We estimated $\E{T_2}$ by $\Pi / 2 \mu$.
For $\Pi$, we used the average pairwise diversity at fourfold degenerate sites in the central region of each chromosome arm.
For $\mu$, we used a genome-wide mutation rate of $3 \times 10^{-9}$ per basepair per generation \autocite{KeightleyEtAl2014}.
These estimates are not precise, but only serve to scale genetic distances to the correct order of magnitude.

@@

\begin{figure}
\centering
\includegraphics[scale=1]{figures/wfpmi_no_recombination.pdf}
\caption{Weighted frequency pointwise mutual information for three coalescent models: the time-homogeneous Kingman coalescent, the Kingman coalescent with exponential growth ($g=4$), and the beta coalescent ($\alpha=1.45$). For all models $n=39$. Growth and $\alpha$ parameters were chosen to generate average SFS with similar Tajima's D. The diagonal, $i=j$, is masked. \label{fig:nonrecombining_pmi}}
\end{figure}

As a first example, \fig{fig:nonrecombining_pmi} shows the $\wfpmi$ for constant-$N$ Kingman coalescent; the Kingman coalescent with exponential growth, $N(t)=N_0 \exp(-g \frac{t}{N_0})$; and a beta coalescent intermediate between the Kingman and Bolthausen-Sznitman coalescents.
Exponential growth with $g=4$ and the beta coalescent with $\alpha=1.45$ both generate similar substantial distortions in the SFS, $D = -0.4$.
However, exponential growth does not qualitatively change the $\fpmi$.
In particular, $\fpmi_{i,j} < 0$ for all $i \neq j$, as in the constant-$N$ Kingman coalescent.
On the other hand, the beta coalescent generates positive $\fpmi$.
The effect of multiple mergers is strongest on $\fpmi$ between high and low minor allele counts.
This justifies binning the SFS and 2-SFS into high-- and low--minor allele count bins as defined above.

To generalize this finding, we computed the binned $\hilopmi$ between singletons and non-singletons ($i_c = 1$) for a range of $g$ and $\alpha$.
\Fig{fig:fpmi_vs_tajimasd} shows that as the population growth rate increases, distorting the SFS, $\hilopmi$ increases relative to the constant-$N$ Kingman, but plateaus at a negative value.
On the other hand, beta coalescents that generate similar distortions in the SFS, generate larger changes in $\hilopmi$, including positive values.
Thus, the 2-SFS, and $\hilopmi$ in particular, are capable of capturing the effects of multiple mergers beyond the distortions in branch lengths.

@@

We have shown in the previous sections that $\fpmi$ between pairs of non-recombining sites can discriminate between the Kingman coalescent with population growth and the beta coalescent.
However, most demographic inference is performed on regions of the genome with non-zero recombination rates.
Therefore, it is important to assess the robustness of our approach to recombination between sites.

To measure the effect of recombination on $\fpmi$, we ran coalescent simulations using a version of the program \texttt{msprime} \cite{KelleherEtAl2016} modified to allow for multiple mergers (See Methods).
In particular, we simulated a model where the sizes of coalescent events are drawn from the distribution specified by the beta coalescent, as in our numerical calculations above.
In this model, marginal genealogies will follow the beta coalescent distribution, and the average SFS will be given by the formula in \cite{BirknerEtAl2013}.
We also simulated data from two models of population growth: exponential growth and a two-epoch piecewise-constant model.

@@

\begin{figure}
\centering
\includegraphics[scale=1]{figures/hilopmi_vs_distance.pdf}
\caption{Pointwise mutual information between high-- and low--minor allele count mutations ($\hilopmi$) versus genetic distance. Each curve represents a different combination of coalescent parameters. The first three rows show coalescent simulations in \texttt{msprime}. Exponential growth rate $g$ varies logarithmically from 0.01 to 8.0 in coalescent time units. Piecewise constant change-points, range from 0.01 to 1 in coalescent time units, and the ratio of ancestral to modern population size ranges from 0.01 to 0.2. The fourth row shows forward-time \texttt{SLiM} simulations with selective sweeps at linked sites (beneficial mutation fraction $f_{\text{sel}} \in \{10^{-6}, 10^{-5}, 10^{-4}, 10^{-3}\}$, see Methods). Lines are colored by the distortion in the SFS, as measured by Tajima's D. The dashed line shows the constant-$N$ Kingman coalescent. \label{fig:hilo_vs_d}}
\end{figure}

\Fig{fig:hilo_vs_d} shows $\hilopmi(d,i_c)$ in both models of growth and the beta coalescent, for a range of $d\,r \E{T_2}$ and three different choices of $i_c$.
Each curve represents a particular parameter combination and coalescent model and is colored by the distortion in the average SFS (Tajima's D), which is independent of the recombination rate.
At low recombination rates, $\hilopmi$ may be greater in growing populations than in constant-$N$ populations (dotted lines), but is always less than zero, consistent with the results for non-recombining sites.
When $d\,r \E{T_2}\geq 1$, $\hilopmi$ may be slightly positive for large $i_c$, but is smaller in growing populations than in the constant-$N$ Kingman model.
In all Kingman models, $\hilopmi$ decays to zero for $d\,r \E{T_2}\gg 1$.

@@

There is one other salient feature of \fig{fig:hilo_vs_d}: $\hilopmi$ does not decay to zero at long distances in the beta coalescent.
This behavior relates to the results of \cite{EldonWakeley2006}, who showed that a model with ``jackpot'' reproductive events can generate infinite-range linkage disequilibrium.
Further, \cite{EldonWakeley2006} showed that different scalings of rates of mutation, pairwise coalescence, multiple merger coalescence, and recombination lead to different behaviors of diversity and linkage disequilibrium.
Our implementation of the beta coalescent model with recombination corresponds to a particular scaling limit where recombination does not have time to decorrelate trees during multiple mergers events, even at infinite genetic distances.
We do not expect this behavior to be universal in multiple mergers coalescents with recombination.

@@

To test this proposition, we performed forward-time simulations of recurring selective sweeps using the software \texttt{SLiM} \autocite{Messer2013}.
In these simulations, we simulated individual chromosomes with homogenous recombination and two types of mutations: neutral and beneficial with a fixed selection coefficient, $s$.
Both types of mutations occurred at random, uniformly distributed along the chromosome.
By varying the recombination rate, neutral and beneficial mutation rates, population size, and selection coefficient, we varied the rate of selective sweeps linked to neutral sites.

The bottom row of \fig{fig:hilo_vs_d} shows the results of these simulations.
For intermediate genetic distances, $d\,r \E{T_2} \sim 1$, the effects of linked sweeps are qualitatively similar to the effects of the beta coalescent.
That is, when sweeps are sufficiently frequent to distort the SFS, as measured by Tajima's D, they also increase $\hilopmi$.
The primary difference between the forward-time sweeps and beta coalescent simulations, is that the distortions caused by sweeps decay to zero for $d\,r \E{T_2} \gg 1$.
This decay is expected because the effect of a single selective sweep on the genealogy is localized around the position of beneficial mutation.

@@

Our results above show that $\fpmi$ and its binned analog, $\hilopmi$, are useful for distinguishing population growth from the effects of multiple mergers, even when population growth generates similar distortions in the SFS.
We therefore propose the following model-checking procedure for demographic inference methods
\begin{enumerate}[label=(\roman*), before=\unskip{: }]
    \item Fit a demographic model to data.
    \item Simulate genealogies under the fitted model (using \texttt{msprime} or other coalescent simulator). Calculate the $\fpmi(d)$ and $\hilopmi(d;i_c)$ predicted by the model.
    \item Calculate $\fpmi(d)$ and $\hilopmi(d;i_c)$ from the data.
    \item Compare true to predicted statistics. Evaluate model fit.
\end{enumerate}
This procedure checks whether the demographic model is consistent with a feature of the data that was not used in fitting the model.
Inconsistency suggests that the inferred $N(t)$ may be an artifact of natural selection, skewed offspring distributions, etc., rather than reflecting the true historical population size.

In this section, we illustrate the procedure outlined above by using genomic diversity data from the \textit{Drosophila melanogaster} DGPG3 panel \cite{LackEtAl2015}.
The DPGP3 data consists of haploid consensus sequences from $\sim$ 200 wild-caught flies from a Zambian population known to be relatively free of cosmopolitan admixture.
Recently, several groups have used the DPGP3 data to estimate the population-size history of \textit{D. melanogaster} \autocite{TerhorstEtAl2017,RagsdaleGutenkunst2017}.
On the other hand, it is widely believed that the genetic diversity of \textit{Drosophila} is strongly shaped by natural selection (e.g., \cite{ElyashivEtAl2016, GarudPetrov2016}).
Thus, this data is a good candidate for demonstrating the utility of $\fpmi$ for assessing coalescent model fit.
